Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups\thanks{The work is partially supported by Siemens Power and Gas. The code is published on GitHub: \href{https://github.com/karhunenloeve/PHIA}{https://github.com/karhunenloeve/PHIA}.}

### 2020-04-02 llncs2e/paper77.tex
typo
typos

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups\thanks{The work is partially supported by Siemens Power and Gas. The code is published on GitHub: \href{https://github.com/karhunenloeve/PHIA}{https://github.com/karhunenloeve/PHIA}.}}
%
\titlerunning{Invertible Resnets up to Homology Equivalence}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an elusive number of scientific disciplines. Nevertheless, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After processing through a layer, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer for an invertible residual net. We consider the data domain as abelian connected matrix group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $n$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

The idea originates from Lie theory. We use invertible architectures, to guarantee our assumptions and apply ascending sorting on the samples to yield permutation invariance or abelianization of the input domain \citemain{ArdizzoneKRK19,NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is capable to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The zeroth homotopy group gives information about connected components which can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so on. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives us a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to predict how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{T}^p \times \mathbb{R}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce to the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training auto encoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

\section{The Manifold Assumption}
\label{sec:manifoldassumption}
The manifold assumption states that a point $x = \{x_1, \cdots, x_d\} \in \mathbb{R}^n$ lies on an embedded manifold $M$, with $\dim M \ll n$. We specify the concept of smooth manifolds. Then we dive into Lie groups concluding with a central theorem on the isomorphism of abelian connected Lie groups to the product space $\mathbb{T}^p \times \mathbb{R}^q$, which turns out useful to estimate the embedding dimension of the manifold.

@@

\subsection{Background: Lie groups}
\label{liegroups}
A Lie group is a $\mathcal{C}^\infty$-manifold $G$ which is equipped with a group structure such that the maps $\mu:(x,y) \mapsto xy, G \times G \rightarrow G$ and $\imath: x \mapsto x^{-1}, G \rightarrow G$ are smooth. In our case $\mu$ corresponds to the matrix multiplication. The neutral element of a group is denoted by $e = e_G$. Each abelian commutative Lie group $G$ of dimension $\dim G = m$ is isomorphic to a particular product space $\mathbb{T}^p \times \mathbb{R}^q$, with $p + q = m$, see Appx. \ref{connectedcommutativeliegroupdecomposition}. Thus, the group can be decomposed into a product space with a relatively simple topological structure. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, so the $q$-torus is the product space $\mathbb{T}^q := \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. Our initial decomposition of the Lie group into a product space can be further simplified into $G \simeq \mathbb{R}^p \times (\mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q)$. Note, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; ||x||_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into a two-dimensional Euclidean space. The homology groups and the embedding dimensions of the product spaces, yield by decomposition of connected commutative Lie groups, are therefore known. Any Euclidean space like $\mathbb{R}^p$ is contractible, i.e. zero-homotopic or homotopy equivalent to a constant map, thus the homology group $H_k(\mathbb{R}^p) = 0$ for $k > 0$ and $H_k(\mathbb{R}^p) = \mathbb{Z}$ for $k = 0$. To illustrate the role of orientation, we calculate the homology groups and illustrate the triangulation of the $1$-sphere, both with coefficients in $\mathbb{R}$ and in $\mathbb{Z}/2\mathbb{Z}$ in Fig. \ref{complexes}. The latter field neglects orientation. Note, that for this field $+1 = -1$.

A tool is needed to estimate homology groups of the manifold on which the data points lie. The method can then be inversely applied to the homology groups to estimate the embedding dimension of the data manifold. We introduce persistent homology with its confidence intervals in order to obtain the best possible estimate of the current state of the art by removing topological noise.

\section{Persistent Homology}
\label{sec:persistenthomology}
In algebraic topology spaces are studied as the composition of triangles, the latter being used as the fundamental component of a large class of spaces. All closed surfaces can be constructed by identifying -- or, less formally, `gluing' -- the edges of triangles \citemain[p.~102]{hatcher2002algebraic,lume}. The $n$-dimensional generalization of triangles are $n$-\emph{simplices}. They are the smallest convex set in Euclidean space with $(n+1)$-points $v_0,\cdots,v_n$, which have no solutions for any system of linear equations in $n$-variables. Thus, they do not lie on any hyperplane with a dimension less than $n$ \citemain[p.~103]{hatcher2002algebraic}. We define the $n$-simplex as follows:
\begin{equation}
    \Delta^n = \left\{ [v_0,\cdots,v_n] \in \mathbb{R}^{n+1} \; \bigg\vert \; \sum_{i=0}^{n} v_i = 1 \; \text{and} \; v_i \geq 0 \; \text{for all} \; i \right\}.
\end{equation}
Removing a vertex from the $\Delta^n$ results in a $(n-1)$-simplex called \emph{face} of $\Delta^n$. The combinatorial description of $n$-simplices is called \emph{simplicial complex}.

>>>>

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups\thanks{The work is partially supported by Siemens Power and Gas. The code is published on GitHub: \href{https://github.com/karhunenloeve/PHIA}{https://github.com/karhunenloeve/PHIA}.}}
%
\titlerunning{Invertible Resnets up to Homology Equivalence}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an elusive number of scientific disciplines. Nevertheless, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After processing through a layer, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer for an invertible residual net. We consider the data domain as abelian connected matrix group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

The idea originates from Lie theory. We use invertible architectures, to guarantee our assumptions and apply ascending sorting on the samples to yield permutation invariance or abelianization of the input domain \citemain{ArdizzoneKRK19,NguyenAK19}. Our interpretation of a neural net is a continuous transformation, transferring invariant structures of the data in the input space to its deepest layer. Thus, it is capable to reconstruct the structure of the original space. This approach corresponds to the concept of homotopy equivalence \citemain{RubinsteinR08,YangY93}, which means that there is a continuous map $f: X \rightarrow Y$ and $g: Y \rightarrow X$, such that $f \circ g$ is homotopic to the identity $\text{Id}_Y$ and $g \circ f$ to $\text{Id}_X$. Thus, the spaces can be deformed continuously into one another. To do so, their homotopy groups must be equal. The zeroth homotopy group gives information about connected components which can be collapsed to a point by continuous functions, the first homotopy group about `holes' and the second about `cavities' and so on. But homotopies are difficult to compute \citemain{klaus2004quick,hatcher2002algebraic}. Homology theory gives us a remedy that guarantees an isomorphism between homotopy groups and homology groups under the conditions of Hurewicz's Theorem \citemain{klaus2004quick}. We use theoretical results from Lie theory and persistent homology theory to predict how many dimensions of embedding are needed to represent the homology groups. Our paper is structured as follows:
\begin{itemize}
    \item In Sect. \ref{sec:manifoldassumption} we discuss the mathematical assumptions and recall the principles of smooth manifolds and Lie theory. In particular, we recall the decomposition of connected commutative Lie groups into $\mathbb{R}^p \times \mathbb{T}^q$.
    \item In Sect. \ref{sec:persistenthomology} we introduce to the theory of persistent homology, define persistence diagrams, persistence landscapes and confidence sets, in order to estimate homology groups despite noisy data. We use Künneth's theorem and derive how to count the homology groups of the input space.
    \item In Sect. \ref{sec:neuralnettheory} we explain the mathematical theory of neural nets and construct invertible deep neural nets to satisfy our assumptions. Resnets, which have a larger state space dimension, offer an architecture with a comparatively small number of parameters to dense nets.
    \item In Sect. \ref{sec:numericalexperiments} we validate our theory training autoencoders on the cifar10 and cifar100 datasets, isolating the innermost layer such that no residual connections bypass it on different embedding dimensions.
\end{itemize}

\section{The Manifold Assumption}
\label{sec:manifoldassumption}
The manifold assumption states that a point $x = \{x_1, \cdots, x_d\} \in \mathbb{R}^n$ lies on an embedded manifold $M$, with $\dim M \ll n$. We specify the concept of smooth manifolds. Then we dive into Lie groups concluding with a central theorem on the isomorphism of abelian connected Lie groups to the product space $\mathbb{R}^p \times \mathbb{T}^q$, which turns out useful to estimate the embedding dimension of the manifold.

@@

\subsection{Background: Lie groups}
\label{liegroups}
A Lie group is a $\mathcal{C}^\infty$-manifold $G$ which is equipped with a group structure such that the maps $\mu:(x,y) \mapsto xy, G \times G \rightarrow G$ and $\imath: x \mapsto x^{-1}, G \rightarrow G$ are smooth. In our case $\mu$ corresponds to the matrix multiplication. The neutral element of a group is denoted by $e = e_G$. Each abelian commutative Lie group $G$ of dimension $\dim G = m$ is isomorphic to a particular product space $\mathbb{R}^p \times \mathbb{T}^q$, with $p + q = m$, see Appx. \ref{connectedcommutativeliegroupdecomposition}. Thus, the group can be decomposed into a product space with a relatively simple topological structure. The $q$-torus $\mathbb{T}^q$ is a surface of revolution. It moves a curve around an axis of rotation. These axes are given by $1$-spheres, so the $q$-torus is the product space $\mathbb{T}^q := \mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q$. Our initial decomposition of the Lie group into a product space can be further simplified into $G \simeq \mathbb{R}^p \times (\mathcal{S}^1_1 \times \cdots \times \mathcal{S}^1_q)$. Note, that the $(n-1)$-sphere is given by $\mathcal{S}^{n-1} := \{x \in \mathbb{R}^n \; \vert \; ||x||_2 = 1\}$. Thus, $\mathcal{S}^1$ can be embedded into a two-dimensional Euclidean space. The homology groups and the embedding dimensions of the product spaces, yield by decomposition of connected commutative Lie groups, are therefore known. Any Euclidean space like $\mathbb{R}^p$ is contractible, i.e. zero-homotopic or homotopy equivalent to a constant map, thus the homology group $H_k(\mathbb{R}^p) = 0$ for $k > 0$ and $H_k(\mathbb{R}^p) = \mathbb{Z}$ for $k = 0$. To illustrate the role of orientation, we calculate the homology groups and illustrate the triangulation of the $1$-sphere, both with coefficients in $\mathbb{R}$ and in $\mathbb{Z}/2\mathbb{Z}$ in Fig. \ref{complexes}. The latter field neglects orientation. Note, that for this field $+1 = -1$.

A tool is needed to estimate homology groups of the manifold on which the data points lie. The method can then be inversely applied to the homology groups to estimate its embedding dimension. We introduce persistent homology with its confidence intervals in order to obtain the best possible estimate of the current state of the art by removing topological noise.

\section{Persistent Homology}
\label{sec:persistenthomology}
In algebraic topology spaces are studied as the composition of triangles, the latter being used as the fundamental component of a large class of spaces. All closed surfaces can be constructed by identifying -- or, less formally, `gluing' -- the edges of triangles \citemain[p.~102]{hatcher2002algebraic,lume}. The $n$-dimensional generalization of triangles are $n$-\emph{simplices}. They are the smallest convex set in Euclidean space with $(n+1)$-points $v_0,\cdots,v_n$, which have no solutions for any system of linear equations in $n$-variables. Thus, they do not lie on any hyperplane with a dimension less than $n$ \citemain[p.~103]{hatcher2002algebraic}. We define the $n$-simplex as follows:
\begin{equation}
    \Delta^n = \left\{ [v_0,\cdots,v_n] \in \mathbb{R}^{n+1} \; \bigg\vert \; \sum_{i=0}^{n} v_i = 1 \; \text{and} \; v_i \geq 0 \; \text{for all} \; i \right\}.
\end{equation}
Removing a vertex from $\Delta^n$ results in a $(n-1)$-simplex called \emph{face} of $\Delta^n$. The combinatorial description of $n$-simplices is called \emph{simplicial complex}.
